---
title: "LDA_final_cz2544"
author: "Chunxiao Zhai"
date: "12/9/2019"
output: word_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
library(lme4)
library(geepack)
library(ggResidpanel)
knitr::opts_chunk$set(echo = TRUE)
```
Objective: Evaluate the efficacy of three different interventions to decrease alcohol assumption in 60 days and to prevent relapse in alcohal dependent patients through a randomized clinical trial.

Abstract(200 words):


1. Checking validity of radomization in terms of gender and disease baseline condition.
```{r read_data}
liquor = read.table(file = "ALCDEP.txt", header = TRUE)
liquor$Treatment = as.factor(liquor$Treatment)
liquor$Gender = as.factor(liquor$Gender)
# liquor$Relapse = as.factor(liquor$Relapse)
discrip = liquor[2:7] %>% skimr::skim() # table 1 

# examin pairwise distribution ballance:
liquor %>% group_by(Treatment, Gender) %>% count() 
liquor %>% ungroup() %>% 
  ggplot(aes(y=ND0,color = Gender))+geom_boxplot()+facet_grid(~Treatment)+
  ggtitle('Number of drinks by gender and intervention, Boxplot')

liquor %>% group_by(Treatment, Gender) %>% select(Relapse) %>% skimr::skim() %>%
  filter(stat=="mean") %>% 
  select(Gender,Treatment,formatted) %>% spread(key = 'Treatment',value = 'formatted') #table 2
```
For the 314 subjects in this study, there are 171 males and 143 females. The randomization assinged 106 to treatment 1, 107 to treatment 2 and 101 to treatment 3. When examine gender distribution in each treatment group, more males than females happened in treatment 2 (62:45) and 3 (57:44). There is no missing value (table 1).
When checking the balance of baseline alcohol dependence, females show significantly lower dependence in all three groups, the distribution of baseline alcohol dependence in 3 treatment groups are similar. The drastic gender difference in disease baseline should be considered in all following analysis (stratify by gender).

The 6 month relapse rate is highest in intervention 1 for both male and female (M:85%, F:80%), lower in intervention 2 (M:45%, F:47%), and lowest in intervention 3 (M:12%, F:16%) (table 2).

2.  Exploratory analysis of treatment effect over time.
```{r EDA}
# put data in long format:
liquor_l = liquor %>%pivot_longer(ND0:ND60, names_to = "time", values_to = "number_of_drinks")
# plot over time in each group
set.seed(2544)
female.1 = liquor %>% filter(Gender==1,Treatment == 1)
female.2 = liquor %>% filter(Gender==1,Treatment == 2)
female.3 = liquor %>% filter(Gender==1,Treatment == 3)
male.1 = liquor %>% filter(Gender==0,Treatment == 1)
male.2 = liquor %>% filter(Gender==0,Treatment == 2)
male.3 = liquor %>% filter(Gender==0,Treatment == 3)

highlight = cbind(sample(female.1$sid,2,replace = FALSE),
                  sample(female.2$sid,2,replace = FALSE),
                  sample(female.3$sid,2,replace = FALSE),
                  sample(male.1$sid,2,replace = FALSE),
                  sample(male.2$sid,2,replace = FALSE),
                  sample(male.3$sid,2,replace = FALSE))
liquor_l %>% 
  ggplot(aes(x=time,y = number_of_drinks, group = sid) )+
  geom_line(color = 'grey')+
  geom_line(aes(color = Gender), data = subset(liquor_l, sid%in%highlight)) +  
  facet_grid(~Treatment)+ggtitle('Number of drinks by gender and intervention, Spaghetti Plot')
```
The treatment effect of each intervention does not differ by gender, but three different inverventions have different patterns. Intervention 2 show continuous decrease during the follow up period. However, intervention 1 and 3 show big drop at 30 day but no change from 30 day to 60 day.

```{r ND0 as factor}
median = liquor %>% group_by(Gender) %>% select(ND0) %>%
  skimr::skim() %>% filter(stat == "p50") %>% select(Gender,formatted) # gender 0 183, gender 1 66
female = liquor %>% filter(Gender==1)%>% mutate(baseline= if_else(ND0>66,"F_Hi","F_Lo"))
male = liquor %>% filter(Gender==0)%>% mutate(baseline= if_else(ND0>183,"M_Hi","M_Lo"))
liquor_l2 = rbind(female,male) %>%
  pivot_longer(ND0:ND60, names_to = "time", values_to = "number_of_drinks") %>% 
  mutate(baseline = as.factor(baseline))

female.1 = female %>% filter(Gender==1,Treatment == 1)
female.2 = female %>% filter(Gender==1,Treatment == 2)
female.3 = female %>% filter(Gender==1,Treatment == 3)
male.1 = male %>% filter(Gender==0,Treatment == 1)
male.2 = male %>% filter(Gender==0,Treatment == 2)
male.3 = male %>% filter(Gender==0,Treatment == 3)
  
highlight_f = cbind(sample(subset(female.1,baseline=="F_Hi")$sid,2,replace = FALSE),
                    sample(subset(female.2,baseline=="F_Hi")$sid,2,replace = FALSE),
                    sample(subset(female.3,baseline=="F_Hi")$sid,2,replace = FALSE),
                    sample(subset(female.1,baseline=="F_Lo")$sid,2,replace = FALSE),
                    sample(subset(female.2,baseline=="F_Lo")$sid,2,replace = FALSE),
                    sample(subset(female.3,baseline=="F_Lo")$sid,2,replace = FALSE))
                  
highlight_m = cbind(sample(subset(male.1,baseline=="M_Hi")$sid,2,replace = FALSE),
                    sample(subset(male.2,baseline=="M_Hi")$sid,2,replace = FALSE),
                    sample(subset(male.3,baseline=="M_Hi")$sid,2,replace =FALSE),
                    sample(subset(male.1,baseline=="M_Lo")$sid,2,replace = FALSE),
                    sample(subset(male.2,baseline=="M_Lo")$sid,2,replace = FALSE),
                    sample(subset(male.3,baseline=="M_Lo")$sid,2,replace = FALSE))

liquor_l2 %>% filter(Gender==1) %>% 
  ggplot(aes(x=time,y = number_of_drinks, group = sid) )+
  geom_line(color = 'grey')+
  geom_line(aes(color = baseline), data = subset(liquor_l2, sid%in%highlight_f)) + 
  facet_grid(~Treatment:baseline)+
  ggtitle('Number of drinks by baseline and intervention in female, Spaghetti Plot')
liquor_l2 %>% filter(Gender==0) %>% 
  ggplot(aes(x=time,y = number_of_drinks, group = sid) )+
  geom_line(color = 'grey')+
  geom_line(aes(color = baseline), data = subset(liquor_l2, sid%in%highlight_m)) + 
  facet_grid(~Treatment:baseline)+
  ggtitle('Number of drinks by baseline and intervention in male, Spaghetti Plot')

female_r = female %>% group_by(Treatment, baseline) %>% select(Relapse) %>% skimr::skim() %>%
  filter(stat=="mean") %>% 
  select(baseline,Treatment,formatted) %>% spread(key = 'Treatment',value = 'formatted')
male_r = male %>% group_by(Treatment, baseline) %>% select(Relapse) %>% skimr::skim() %>%
  filter(stat=="mean") %>% 
  select(baseline,Treatment,formatted) %>% spread(key = 'Treatment',value = 'formatted')
```

When split the data by median baseline alcohol dependence in each gender group, the number of drinks of high baseline subjects in both genders drop faster than low baseline subjects in first 30 days of intervention, to the extent that number of drinks in the first 30 days are similar for subjects in the same treatment group regardless of baseline dependence. The change from number of drinks in first 30 days (0 to 30), to second 30 days (30 to 60)
for both bathline levels are similar in male and female.
6 month relapse rates are not affected by baseline alcohol dependence level except for male patients in intervention 3, where subjects with high baseline level have lower relapse rate. Potiential influencial points need to be examined.

3. Model building for treatment effect on number of drinks and 6 month relapse rate.
```{r ND_mdl}
# start with full model to include all potential interactions
# buid mixed effect model to allow subject specific features

glm1 = glmer(number_of_drinks ~ Treatment*baseline*time+(1|sid),
             data=liquor_l2,family=poisson(link = "log")) 
# baseline contain gender info
summary(glm1) 
# baseline significant, treatment:time significant, baselline:time significant
# 3-way interaction not significant
# stratify by gender:
glm2.1 = glmer(number_of_drinks ~ Treatment*time+baseline*time+(1|sid),
             data=subset(liquor_l2,Gender==1),family=poisson(link = "log")) 
summary(glm2.1)
# Create an interactive panel of the default diagnostic plots
resid_interact(glm2.1) #fit good, sensitivity analysis check sid 301, 289
resid_xpanel(glm2.1)

# remove sid 301, 289
data_301 = liquor_l2 %>% filter(sid!=301)
data_289 = liquor_l2 %>% filter(sid!=289)
glm2.1.1 = glmer(number_of_drinks ~ Treatment*time+baseline*time+(1|sid),
             data=subset(data_301,Gender==1),family=poisson(link = "log")) 
summary(glm2.1.1) # 301 does not change conclusion
glm2.1.2 = glmer(number_of_drinks ~ Treatment*time+baseline*time+(1|sid),
             data=subset(data_289,Gender==1),family=poisson(link = "log")) 
summary(glm2.1.2) # 289 does not change conclusion


glm2.2 = glmer(number_of_drinks ~ Treatment*time+baseline*time+(1|sid),
             data=subset(liquor_l2,Gender==0),family=poisson(link = "log")) 
summary(glm2.2)
resid_interact(glm2.2) #fit good
resid_xpanel(glm2.2)

#for 2 genders' results similer, try only gender no baseline model(treatment:gender not significant):
glm2.3 = glmer(number_of_drinks ~ Treatment*time+Gender+(1|sid),
             data=liquor_l,family=poisson(link = "log")) 
summary(glm2.3)
resid_interact(glm2.3)
```
For male patients and female patients have very different baseline levels, analysis of treatment effect is only compared in each gender stratum.
Fit linear mixed effect model in each gender, baseline alcohol dependence for a typical female patient in female high baseline group is 76.67, in female low baseline group is 58.26. A typical female subject in intervention group 3 has the biggest decrease in alcohal dependence at both follow-ups. A typical subject with high baseline level has bigger decrease of alcohol dependence at both followups when comepare to a typical subject in low baseline level group.
In male patients, similarly, baseline alcohol dependence for a typical patient in male high baseline group is 204.41, in male low baseline group is 167.72. A typical male subject in intervention group 3 has the biggest decrease in alcohal dependence at both follow-ups. A typical subject with high baseline level has bigger decrease of alcohol dependence at both followups when comepare to a typical subject in low baseline level group.

When baseline level is not considered, treatment 3 remain the best for both genders.

```{r rela_mdl}
# start with full model to include all potential interactions
# in female:
lme1 = glmer(Relapse ~ Treatment*baseline+(1|sid),data=female,id = sid,family=binomial(link = "logit"))
summary(lme1)
resid_interact(lme1)
# in male:
lme2 = glmer(Relapse ~ Treatment*baseline+(1|sid),data=male,family=binomial(link = "logit")) # baseline contain gender info
summary(lme2)
resid_interact(lme2)

# both gender show treatment 3 significantly better, no treatment:baseline interaction
# combine genders
lme3 = glmer(Relapse ~ Treatment*Gender+(1|sid),data=liquor,family=binomial(link = "logit")) # baseline contain gender info
summary(lme3)
resid_interact(lme3)
```
When compare two treatments

4. Results

5. Conclusion

6. Caveats 
i)model limitation
ii)unexpected results
iii)suggestion for future studies

7. Discussion
